### 2019-07-18 manuscript/response.tex

\medskip
Thank you very much for your close reading of our manuscript and supportive comments. We appreciate your constructive suggestions, each of which we address below. You will find a great difference in the number of models that we have fitted for the revision based on reviewer suggestions (originally 12, now 29, with a special supplementary figure to enable easy visualization). There are two main reasons for this sizable increase.

The first reason is additional work to address a key concern from all referees, about the transition symmetry among the hidden states. 
Briefly, we found that is relevant to have asymmetric hidden state rates and, in our case, it increased the differences among net diversification rates. 
We added and altered the text to reflect these additional analyses.

The second reason is related to a question we mulled over after submitting the first version of this article. Specifically, we wondered whether increasing the state space from two- to three-states in our models of the diversification process explains the data better (and how they ought to be best compared). 
This  trait coding question could not be addressed directly by model comparison statistics because the two-state and three-state models need different input (i.e. (0,1) versus (0,1,2)). 
The species and the tree are identical, however, and our intuition indicated that they should carry the same biological evidence despite a particular coding input. 
During this revision, we figured a way to recode the three-state models to make them equivalent to the two-state models (the new ``lumped" models). 
This equivalency enables comparison of two-state and three-state models, and assessment of their relative explanatory value after adding key trait information and states in diversification analyses. 
We also revised the original dataset (thanks to the suggestion of Reviewer 1) and we were able to correct discrepancies with a previous dataset, as well as add more data (now 651 taxa).

W hope the changes and the new models are satisfactory to the editorial staff and the reviewers.

@@

\medskip
Sincerely,

Rosana Zenil-Ferguson, Gordon Burleigh, Boris Igi\'c, Itay Mayrose, and Emma Goldberg.

### 2019-07-19 manuscript/discussion.tex
- Final ms changes pass-through, mainly spaces, commas, spelling, etc. - Leaving FIXME marks for galley proofs. - Will update archive once more with submission version.

%R- let's not get into the weeds of what is a hidden state (I remove the use of trait  for state because a state is what the model receives). Jeremy does a good job discussing this in his paper so no need.
%     but don't be circular, test in different clade
% alpha is always low: new supp fig showing how A & B are inferred on the tree?  at least for ID/CD/CP+A/B model - it is low because not enough transitions and mostly a different process towards the tips-who knows what is happening there?
%Here Rosana adds the value of adding traits with the lumped models
As more trait information becomes available for macroevolutionary studies, it is not only important to recognize the role of hidden states as a part of a general modeling approach, but also to question whether adding more traits to diversification studies is justified.  More generally, we find that lumped models are useful for assessing the value of adding additional traits to already complex diversification models(\cref{figure:lumped,table:lumped}).
%R-removed this: %In our analyses, we find that the transition rates between and within hidden traits did not affect the direction of the net diversification differences. 
%We find that it is important to allow for the hidden state transition rates to be asymmetrical, because those models tend to provide a better fit than models that fix the hidden state transition rates to be equal (\cref{suppfigure:asymmetric,supptable:asymmetry}).

Although we fit an extensive set of models in order to relax a variety of assumptions, we did not explore the process of trait change in conjunction with speciation.
That is, our models all assume anagenetic trait evolution and ignore cladogenetic shifts.
Anagenetic and cladogenetic changes can be separated with phylogenetic models \citep{mayrose_2011, goldberg_2012, magnuson-ford_2012}. 
These have been applied to estimate the relative contribution of anagenetic and cladogenetic shifts in breeding system \citep{goldberg_2012} and polyploidization \citep{zhan_2016, freyman_2019}.
We did not explore cladogenetic trait change because they would greatly increase in state space and model number of our already complex and extensive modeling framework. %IM: #29 and GB suggested that chance
Although \citet{goldberg_2012} found that allowing cladogenetic changes did not substantially affect inference of net diversification rates associated with breeding system, future work could test whether this process affects diversification rate estimates with the more complex state and parameter spaces of our other models.

>>>>

%R- let's not get into the weeds of what is a hidden state (I remove the use of trait  for state because a state is what the model receives). Jeremy does a good job discussing this in his paper so no need.
%     but don't be circular, test in different clade
% alpha is always low: new supp fig showing how A & B are inferred on the tree?  at least for ID/CD/CP+A/B model - it is low because not enough transitions and mostly a different process towards the tips-who knows what is happening there?
%Here Rosana adds the value of adding traits with the lumped models
As more trait information becomes available for macroevolutionary studies, it is not only important to recognize the role of hidden states as a part of a general modeling approach, but also to question whether adding more traits to diversification studies is justified.  
More generally, we find that lumped models are useful for assessing the value of adding additional traits to already complex diversification models (\cref{figure:lumped,table:lumped}).
%R-removed this: %In our analyses, we find that the transition rates between and within hidden traits did not affect the direction of the net diversification differences. 
%We find that it is important to allow for the hidden state transition rates to be asymmetrical, because those models tend to provide a better fit than models that fix the hidden state transition rates to be equal (\cref{suppfigure:asymmetric,supptable:asymmetry}).

Although we fit an extensive set of models in order to relax a variety of assumptions, we did not explore the process of trait change in conjunction with speciation.
That is, our models all assume anagenetic trait evolution and ignore cladogenetic shifts.
Anagenetic and cladogenetic changes can be separated with phylogenetic models \citep{mayrose_2011, goldberg_2012, magnuson-ford_2012}. 
These have been applied to estimate the relative contribution of anagenetic and cladogenetic shifts in breeding system \citep{goldberg_2012} and polyploidization \citep{zhan_2016, freyman_2019}.
We did not explore cladogenetic trait change because of the requisite increase in parameter space and model number, outside the scope of our already complex and extensive modeling framework. %B: restored IM:#29 and removed "they" pointing to "trait change."
Although \citet{goldberg_2012} found that allowing cladogenetic changes did not substantially affect inference of net diversification rates associated with breeding system, future work could test whether this process affects diversification rate estimates with the more complex state and parameter spaces of our other models.

### 2019-07-19 manuscript/introduction.tex
- Final ms changes pass-through, mainly spaces, commas, spelling, etc. - Leaving FIXME marks for galley proofs. - Will update archive once more with submission version.

Polyploidy and self-fertilization are widely thought to be associated \citep{stebbins1950}.
Whole genome duplication may facilitate the transition to selfing by masking inbreeding depression, or self-fertilization may facilitate polyploidy establishment by by avoiding the lower fitness that characterize early-forming polyploids \citep{levin_1975,ramsey_1998, barringer2007, barrett2008, husband2008}.%IM suggestion and Gordon suggestion
Additionally, in RNase-based gametophytic SI systems, polyploidization directly causes the loss of SI \citep{stout1942, lewis1947}.
%B: disclaimer: ceteris paribus statement omitted; polyploidization ordinarily causes SC because of the formation of heterodiploid pollen. On the background of deletion of S-locus, or with heterologous chromosome homozygosity at the S-locus may not do so. These, however, appear to be sufficiently rare that they are recovered perhaps twice in dicots: once (or twice) naturally, in sour cherry and its close relative (Prunus), and once in a experimentally induced and selected synthetic pentaploid Solanum (which has not been replicated). Strictly speaking, this is a feature of non-self recognition systems of the kind found in RNase-based GSI.
In these systems, SI occurs because haploid self-pollen grains, with one S-allele at the locus controlling the SI response, are unable to detoxify the S-RNase produced by the same S-allele in the style \citep{kubo2010}.
The unreduced pollen of diploids, however, can contain two S-alleles expressed in pollen, which jointly gain the ability to detoxify the S-RNases produced by any maternal genotype \citep{entani1999, tsukamoto2005, kubo2010}.
Initial mutant individuals with pollen containing doubled haploid genomes are consequently capable of self-fertilization, with exceedingly few exceptions \citep{hauck_2002, nunes_2006}.
RNase-based SI is regarded as ancestral in eudicots \citep{igic_2001,steinbachs_2002}, and it is expressed in all SI species of Solanaceae examined to date.
The absence of SI polyploids in this family yields a strong correlation between ploidy and breeding system \citep{robertson_2011}.

We address two macroevolutionary questions about the correlated evolution of ploidy and breeding system.
First, we investigate their joint influence on rates of speciation and extinction.
Each character alone has been shown to be associated with differential lineage diversification, but if their effects are not additive, studying them separately may not reveal their combined effect.
Second, we examine the order of transitions in the two characters.
Evolution commonly proceeds from diploid to polyploid, and from SI to SC states, but there are two paths by which diploid SI lineages can eventually become SC polyploids.
Loss of SI in diploids could be directly caused by polyploidization (as explained above, for RNase-based SI systems), resulting in a one-step pathway to SC polyploids. 
Alternatively, SI diploids could first transition to SC without an increase in ploidy, and subsequently undergo polyploidization, resulting in a two-step pathway to SC polyploids.
\citet{robertson_2011} compared the contributions of these two paths, finding that evolution from SI diploids to SC polyploids is more likely to proceed via the one-step pathway over short timescales, but via the two-step pathway over long timescales.
They considered only transitions among the states, however, and we investigate whether these results hold true when allowing for differences in lineage diversification.

In the present study, we employ an extended framework of state speciation and extinction models, which simultaneously model transitions between the discrete states of a trait and different rates of speciation and extinction associated with each of those states \citep[`SSE' models;][]{maddison_2007, fitzjohn_2012}.
We start by fitting binary state speciation and extinction models to ploidy and breeding system independently \citep{maddison_2007}.
We follow by fitting models that incorporate hidden states, which reduce the chance that the effect of the focal trait (ploidy or breeding system) on diversification is found to be significant when in reality, it is simply background heterogeneity in the diversification process what produces the diversification patterns in the phylogeny \citep{beaulieu_2016}. %IM: added a little more explanation
We compare the proposed models against their character independent counterparts \citep{beaulieu_2016} to investigate whether ploidy or breeding system are actually affect the diversification process.
Next, we model ploidy and breeding system jointly to assess their combined influence on diversification, with or without an additional hidden state.
Using the ploidy and breeding system model without hidden states, we quantify the relative contributions of the two pathways from SI diploids to SC polyploids.
We also aggregate states within these joint models of ploidy and breeding system in order to test whether increasing complexity from one trait to two traits significantly improves our understanding of the diversification process.
Furthermore, we extend all the models involving ploidy to investigate the potential effects of including diploidization.
% And finally, we explore models of diversification using asymmetric rates between and within hidden states to investigate the sensitivity of the statistical inference to model assumptions.
Our results highlight the importance of considering non-additive effects of traits on net diversification rates  under the presence of unobserved factors, in order to detect strong biologically-driven processes dictating the diversification patterns.

>>>>

Polyploidy and self-fertilization are widely thought to be associated \citep{stebbins1950}.
Whole genome duplication may facilitate the transition to selfing by masking inbreeding depression, or self-fertilization may facilitate establishment of early-forming polyploids by avoiding the lower fitness \citep{levin_1975,ramsey_1998, barringer2007, barrett2008, husband2008}.%IM suggestion and Gordon suggestion %B: reworded because polyploidy is a process, so its establishment is not the issue.
Additionally, in RNase-based gametophytic SI systems, polyploidization directly causes the loss of SI \citep{stout1942, lewis1947}.
%B: disclaimer: ceteris paribus statement omitted; polyploidization ordinarily causes SC because of the formation of heterodiploid pollen. On the background of deletion of S-locus, or with heterologous chromosome homozygosity at the S-locus may not do so. These, however, appear to be sufficiently rare that they are recovered perhaps twice in dicots: once (or twice) naturally, in sour cherry and its close relative (Prunus), and once in a experimentally induced and selected synthetic pentaploid Solanum (which has not been replicated). Strictly speaking, this is a feature of non-self recognition systems of the kind found in RNase-based GSI.
In these systems, SI occurs because haploid self-pollen grains, with one S-allele at the locus controlling the SI response, are unable to detoxify the S-RNase produced by the same S-allele in the style \citep{kubo2010}.
The unreduced pollen of diploids, however, can contain two S-alleles expressed in pollen, which jointly gain the ability to detoxify the S-RNases produced by any maternal genotype \citep{entani1999, tsukamoto2005, kubo2010}.
Initial mutant individuals with pollen containing doubled haploid genomes are consequently capable of self-fertilization, with exceedingly few exceptions \citep{hauck_2002, nunes_2006}.
RNase-based SI is regarded as ancestral in eudicots \citep{igic_2001,steinbachs_2002}, and it is expressed in all SI species of Solanaceae examined to date.
The absence of SI polyploids in this family yields a strong correlation between ploidy and breeding system \citep{robertson_2011}.

We address two macroevolutionary questions about the correlated evolution of ploidy and breeding system.
First, we investigate their joint influence on rates of speciation and extinction.
Each character alone is thought to be associated with differential lineage diversification but, if their effects are not additive, studying them separately may not reveal their combined effect.
Second, we examine the order of transitions in the two characters.
Evolution commonly proceeds from diploid to polyploid, and from SI to SC states, but there are two paths by which diploid SI lineages can eventually become SC polyploids.
Loss of SI in diploids could be directly caused by polyploidization (as explained above, for RNase-based SI systems), resulting in a one-step pathway to SC polyploids. 
Alternatively, SI diploids could first transition to SC without an increase in ploidy, and subsequently undergo polyploidization, resulting in a two-step pathway to SC polyploids.
\citet{robertson_2011} compared the contributions of these two paths, finding that evolution from SI diploids to SC polyploids is more likely to proceed via the one-step pathway over short timescales, but via the two-step pathway over long timescales.
They considered only transitions among the states, however, and we investigate whether these results hold true when allowing for differences in lineage diversification.

In the present study, we employ an extended framework of state speciation and extinction models, which simultaneously model transitions between the discrete states of a trait and different rates of speciation and extinction associated with each of those states \citep[`SSE' models;][]{maddison_2007, fitzjohn_2012}.
We start by fitting binary state speciation and extinction models to ploidy and breeding system independently \citep{maddison_2007}.
We follow by fitting models that incorporate hidden states, which reduce the chance that the effect of the focal trait (ploidy or breeding system) on diversification is found to be significant when in reality, it may simply represent background heterogeneity in the diversification process that produces the phylogenetic patterns \citep{beaulieu_2016}. %IM: added a little more explanation
We compare the proposed models against their character-independent counterparts \citep{beaulieu_2016} to investigate whether ploidy or breeding system affect the diversification process.
Next, we model ploidy and breeding system jointly to assess their combined influence on diversification, with or without an additional hidden character. %B: additional state implies that we would introduce another state, say 'C' to 'A' and 'B'.
Using the ploidy and breeding system model without hidden states, we quantify the relative contributions of the two pathways from SI diploids to SC polyploids.
We also aggregate states within these joint models of ploidy and breeding system in order to test whether increasing complexity from one trait to two traits significantly improves our understanding of the diversification process.
Furthermore, we extend all the models involving ploidy to investigate the potential effects of including diploidization.
% And finally, we explore models of diversification using asymmetric rates between and within hidden states to investigate the sensitivity of the statistical inference to model assumptions.
Our results highlight the importance of considering non-additive effects of traits on net diversification rates  under the presence of unobserved factors, in order to detect strong biologically-driven processes dictating the diversification patterns.

### 2019-07-19 manuscript/main.tex
- Final ms changes pass-through, mainly spaces, commas, spelling, etc. - Leaving FIXME marks for galley proofs. - Will update archive once more with submission version.

%--------------------------------------------------
\section{Acknowledgements}
%--------------------------------------------------
Special thanks to Bob Thomson for donating computational resources that made possible testing the convergence of the models. 
We also thank Sergei Tarasov for clarifying lumpability concepts, and Carrie Tribble for troubleshooting some ancestral estimate figures. 
The computing resources were provided by the Minnesota Supercomputing Institute (MSI) at the University of Minnesota.
This work was supported by the National Science Foundation grant DEB-1655478 (to EEG), DEB-1655692 and NESCent sabbatical scholar award EF-0905606 (to BI), and the Israel Science Foundation 961/17 (to IM). %IM: #32
%B: new:
We thank Dan Schoen and three anonymous reviewers for providing uncommonly thoughtful and helpful reviews, as well as Spencer Barrett for continually inspiring our work.

>>>>

%--------------------------------------------------
\section{Acknowledgements}
%--------------------------------------------------
Special thanks to Bob Thomson for donating computational resources that made possible testing the convergence of the models. 
We also thank Sergei Tarasov for clarifying lumpability concepts, and Carrie Tribble for troubleshooting some ancestral estimate figures. 
The computing resources were provided by the Minnesota Supercomputing Institute (MSI) at the University of Minnesota.
This work was supported by the National Science Foundation grants DEB-1655478 (to EEG and IM), NSF DEB-1655692 \& NESCent sabbatical scholar award EF-0905606 (to BI), and the Israel Science Foundation 961/17 (to IM). %IM: #32
We thank Dan Schoen and three anonymous reviewers for providing uncommonly thoughtful and helpful reviews, as well as Spencer Barrett for continually inspiring our work.
